---
title: "PEtab2dMod"
author: "Svenja Kemmer and Marcus Rosenblatt"
date: "March 2020"
output:
  html_document: default
  pdf_document: default
---

[PEtab](https://github.com/PEtab-dev/PEtab/) is a data format thought for the straight forward exchange of models between tools with regard to parameter estimation. It covers SBML models along with their data, observable, parameter and condition specifications. **PEtab2dMod** is a tool to import such PEtab files in dMod. Basic functions are provided for import, visualization and further processing. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 3.5)

```

## Installation

The usage of PEtab2dMod requires libSBML. The libSBML R source package can be downloaded as tar.gz file provided by [SBML](https://sourceforge.net/projects/sbml/files/libsbml/5.18.0/stable/R%20interface/) and installed as R package. PEtab2Mod is delivered along with the current version of dMod?! 

## Loading a model

In order to use PEtab2dMod and import a model, data, observable, condition and parameter files have to meet the [requirements](https://github.com/PEtab-dev/PEtab/blob/master/doc/documentation_data_format.md) specified by PEtab.

As a first step, we have to load dMod, which includes all necessary PEtab2dMod functions as well as a collection of Benchmark models.  


```{r, message=FALSE}

library(dMod)
library(PEtab2dmod) # hopefully not necessary anymore

```

Now import and plotting functions are available. As an example we will import and plot [Boehm_JProteomeRes2014](https://pubs.acs.org/doi/abs/10.1021/pr5006923), which is a STAT5 activation model in BaF3-EpoR cells.


```{r, eval=TRUE}

importPEtabSBML(modelname = "Boehm_JProteomeRes2014", path2model = "../BenchmarkModels/") 

```
```{r}

plotPEtabSBML(name%in%names(observables))

```

**Happy PEtab import!!**
