---
title: "List objects in dMod"
author: "Daniel Kaschek"
date: "December 21, 2018"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 3.5)
```

dMod provides several list objects, i.e., objects which are organized as lists but carry their own class. This vignette is about the typical dMod list objects

* `datalist`
* `prdlist`

and what they do or do not have in common.


```{r, message=FALSE}

library(dMod)
library(dplyr)
setwd(tempdir())

```

## `datalist`

Data lists are used to represent the data in dMod. Data lists are most conveniently produced using the `as.datalist()` function on data frames:

```{r}

myframe <- data.frame(
  # required columns
  name = "A",
  time = 1:5,
  value = c(exp(-0.1 * 1:5), exp(-0.3 * 1:5)),
  sigma = 0.05,
  # additional columns to specify the data records
  treatment = "TRTno",
  inhibitor = rep(c("INHno", "INHx"), each = 5)
)

mydata <- as.datalist(myframe)

mydata

```

All non-standard columns are used to create a `condition` identifier. A separate data frame is returned for each condition, structured as a list. Because the non-standard columns describing the data records could potentially be used as covariates for model parameters, the mapping between conditions and covariates can be accessed via

```{r}
covariates(mydata)
```

Data lists can be

* concatenated using the `c()` function or the `+` operator,
* plotted with `plot()` or `plotData()`, and
* converted back to a data frame (including the additional covariate columns) by `as.data.frame()`.

Whenever parameters are estimated from data with dMod, a valid data list needs to be created.

## `prdlist`

Prediction lists are the counterpart of data lists. They are returned by prediction functions. The prediction function returns a separate matrix with the simulation result for each condition, structured as a list.

Prediction lists can be

* plotted with `plot()` or `plotPrediction()`,
* converted to a data frame with `as.data.frame(x, data = NULL, errfn = NULL)`. Optionally, a data list can be provided from where the covariate table is used to produce extra covariate columns in the output data frame. If an error model ist provided (`errfn` is formally an observation function), it is used to impute the `sigma` column.
* Model parameters used to compute the prediction can be extracted with `getParameters()`. 

## Additional lists in dMod

There are some other list structures in dMod which fit a bit less into the schematics presented above. These are `eqnlist` (model representation), `objlist` (list with value, gradient and Hessian returned by an objective function) and `parlist` (a list of parameter fits as returned by `mstrust()`). In the following, we will look a bit close into `eqnlist` and `parlist`.

### `eqnlist`

Equation lists (`eqnlist`) are used to store differential equation models in a list containing the stoichiometric matrix, the states, the rate expressions, volume expressions and a description field:

```{r}

reactions <- eqnlist() %>% 
  addReaction("A", "B", "k_on*A", "Forward reaction") %>% 
  addReaction("B", "A", "k_off*B", "Backward reaction")

unclass(reactions)

```

* Models represented by separate equation lists can be combined, using `c()`. 
* Equation lists can be written to disk using `write()`.
* They can be subsetted using `subset()`.
* They can be converted to differential equations by `as.eqnvec()`.
* Use `as.data.frame()` to extract the stoichiometric matrix and rate expressions as data frame.
* Parameters can be extracted from the rate expressions using `getParameters()`.
* Use `getReactions()` to get the reactions as reaction table.
* `getFluxes()` returns the reaction fluxes from and to every state.

### `parlist`

Parlists are defined by the `mstrust()` output: a list of all the fits. The most important method to run on `parlist` objects is 

* `as.parframe()` which extracts the fitted parameter values from the fits and structures them in a nice data frame and, subsequently,
* `as.parvec()` which extracts the first parameter vector (or the one with the lowest objective value) from the parframe.


**Happy listing!!**
